---
title: "SpecialSchools"
author: "Alexandra Pasternakova"
date: "5/1/2022"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
---

```{r setup, include=TRUE, results="hide", echo=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
               cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r Load libraries}
library(Rcpp)
library(fasterize)
library(raster)
library(sf)
library(dplyr)
library(tidyverse)
library(mapboxapi)
library(leaflet)
library(mapview)
library(RColorBrewer)
library(viridis)
```

```{r wd, include=FALSE, results="hide", echo=FALSE}
# set working directory
setwd("/Final_Project")
```

```{r mapbox-auth, include = FALSE, results="hide", echo=FALSE}
# mapbox token
# my_token <- ""
# mb_access_token(my_token, install = TRUE)
```

## Load data
```{r Load schools data}
filename = "../data/w_data/schools_new_cl.csv"
sch <- read_csv(file= filename, col_names = TRUE)

```


```{r mapbox-map}

mapbox_map <- leaflet() %>%
  addMapboxTiles(style_id = "streets-v11",
                 username = "mapbox") 

```

# Map of all the schools
```{r Plot schools}
# check if our columns are correct type
sapply(sch, class)

# map with all the schools
leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(sch$lng, sch$lat, popup = sch$label)

```


```{r Create sf object}
# Convert to an sf object
sch_sf <- st_as_sf(sch, coords = c("lng", "lat"), crs = 4326) %>% 
        st_transform(crs = 25832)
```


# Walking to school 5km
```{r Walking distance isochrones}
# create isochrones for 3 selected distances
walking_isos <- mb_isochrone(sch_sf,
                             profile = "walking",
                             distance = c(2000, 3000, 5000),
                             id = "label")


# save the isos object
write_sf(walking_isos, "../data/w_data/walk_isos4326.shp")

# set my colors
colors1 <- viridisLite::viridis(3)

# map of the walking isocgrones color-coded based on different distance
map_walk <- mapbox_map  %>%
  addPolygons(data =walking_isos,
             popup = ~id,
             color = rev(colors1),
             fillColor = rev(colors1),
             fillOpacity = 0.5,
             opacity = 0.5,
             weight = 0.2) %>%
  addLegend(position = "topright",
           labels =  c('2km', '3km', '5km'),
           colors = colors1,
           title = "Walking to school")
# 
map_walk # this map is great, however, the overlapping isochrones in some areas seem quite messy so I will create a raster for better visualization
```


```{r Create raster and plot again, warning=FALSE}
# transform projection
walking_isos_proj <- st_transform(walking_isos, 25832)

# make raster with 100m resolution (it would be nicer with better res, however, my computer can't handle that)
template <- raster(walking_isos_proj, resolution = 100)
# assign minimum overlapping value from isochrones to each grid cell
iso_surface <- fasterize(walking_isos_proj, template, field = "distance", fun = "min")

# make a map
map_walk2 <- mapbox_map %>% 
  addRasterImage(iso_surface,
                 colors = colors1,
                 opacity = 0.5) %>%
  addLegend(position = "topright",
           labels =  c('2', '3', '5'),
           colors = colors1,
           title = "Walking reach (km)")

map_walk2 # now we can see the areas accessible without the mess and we can also see the map underneath

# save the raster layer
writeRaster(iso_surface, '../data/raster_data/w_iso_surface100.tif', format = 'GTiff', overwrite = TRUE)

```


## Area outside of 5km walk - whole Denmark
```{r Load Denmark data}
# loading data for Denmark and transforming to same epsg
DK <- getData("GADM", country = "DNK", level = 0)
DK_tf <- st_transform(st_as_sf(DK), crs = 32618)
DK_tf <- st_transform(DK_tf, 25832)
st_crs(DK_tf)

# check with a plot
plot(st_geometry(DK_tf), col = "grey", border = "white")

```


```{r Calculating the area not covered, warning=FALSE}
# some polygons are overlapping and we would get an error trying to do st_union, we can fix that by first projecting 
# them into 2D coordinates
walking_tf <- st_transform(walking_isos, 25832)

# merging all the polygons
walking_un_tf <- st_union(walking_tf)

# calculating the areas not covered by walking polygons
difference <- st_difference(DK_tf, walking_un_tf)
# make sure our crs is the same and nothing weird happened
st_crs(difference)

# plot to make sure everything looks good
plot(st_geometry(difference), col = "grey", border = "white")

# st_area 
st_area(DK_tf) #area of the whole Denmark
st_area(difference) #area which can't reach a special school within 5km walk
# same as above but divided so we can see km for easier comprehension
st_area(difference)/1000000 # change to km^2
st_area(difference)/st_area(DK_tf) # get percentage

```

40186518973 m^2 is 40186.52km^2
The area of Denmark not able to reach a Special school in 5km walk is 40 186.523km^2 which is 93,19%.


# Biking 5km
```{r}
biking_isos <- mb_isochrone(sch_sf,
                             profile = "cycling",
                             distance = c(2000, 3000, 5000),
                             id = "label")


# map of the walking isocgrones color-coded based on different distance
map_bike <- mapbox_map  %>%
  addPolygons(data =biking_isos,
             popup = ~id,
             color = rev(colors1),
             fillColor = rev(colors1),
             fillOpacity = 0.5,
             opacity = 0.5,
             weight = 0.2) %>%
  addLegend(position = "topright",
           labels =  c('2km', '3km', '5km'),
           colors = colors1,
           title = "Biking to school")
#map 
map_bike

```


```{r}
# change to sf_object
biking_tf <- st_transform(biking_isos, 25832)

# merging all the polygons
biking_un_tf <- st_union(biking_tf)

# calculating the areas not covered by walking polygons
difference <- st_difference(DK_tf, biking_un_tf)
# make sure our crs is the same and nothing weird happened
st_crs(difference)

# plot to make sure everything looks good
plot(st_geometry(difference), col = "grey", border = "white")

# st_area 
st_area(DK_tf) #area of the whole Denmark
st_area(difference) #area which can't reach a special school within 5km walk
# same as above but divided so we can see km for easier comprehension
st_area(difference)/1000000 # change to km^2
st_area(difference)/st_area(DK_tf) # get percentage

```

40483614510 m^2 is 40483.61km^2
The area of Denmark not able to reach a Special school in 5km bike ride is 40 483.61km^2 which is 93.88%.Biking covers sligtly less area but the difference is very small. 


# Driving to school 30 minutes
```{r Driving time isochrones}
# create isochrones of 20 minute rides 
driving_isos <- mb_isochrone(
  sch_sf,
  profile = "driving",
  time = c(20,30),
  id = "label"
)

# save the object
write_sf(driving_isos, "../data/w_data/drive_isos4326.shp")

# set my color
colors2 <- c("#C70039", "#FFC300")

# map
map_drive <- mapbox_map  %>%
  addPolygons(data =driving_isos,
             popup = ~id,
             color = rev(colors2),
             fillColor = rev(colors2),
             fillOpacity = 0.5,
             opacity = 0.5,
             weight = 0.2) %>%
  addLegend(position = "topright",
           labels =  c("20","30"),
           colors = colors2,
           title = "Drive reach (minutes)")
# 
map_drive


```

```{r Create raster and plot driving}
# transform projection
driving_isos_proj <- st_transform(driving_isos, 25832)

# make raster with 100m resolution
template <- raster(driving_isos_proj, resolution = 100)
# assign minimum overlapping value from isochrones to each grid cell
iso_surface2 <- fasterize(driving_isos_proj, template, field = "time", fun = "min")

map_drive2 <- mapbox_map %>% 
  addRasterImage(iso_surface2,
                 colors = colors2,
                 opacity = 0.5) %>%
  addLegend(position = "topright",
           labels =  c("20","30"),
           colors = colors2,
           title = "Drive reach (minutes)")

map_drive2 

# save the raster layer
writeRaster(iso_surface2, '../data/raster_data/d_iso_surface100.tif', format = 'GTiff', overwrite = TRUE)

```

## Area outside of 30 minute drive - whole Denmark
```{r Calculate area for driving, warning=FALSE}
# projection
driving_tf <- st_transform(driving_isos, 25832)

# merging all the polygons
driving_un_tf <- st_union(driving_tf)

# calculating the areas not covered by driving polygons
difference2 <- st_difference(DK_tf, driving_un_tf)
# make sure our crs is the same
st_crs(difference2)

# plot
plot(st_geometry(difference2), col = "grey", border = "white")

# st_area 
st_area(DK_tf) #area of the whole Denmark
st_area(difference2) #area which can't reach a special school in 30min ride
st_area(difference2)/1000000 # change to km^2
st_area(difference2)/st_area(DK_tf) # get percentage

```

6557964856 m^2 is 6 557.965km^2
The area of Denmark not able to reach a Special school in 30 minute drive is 16 524,78 km^2 which is 15,21%.


# Covering all of DK
```{r Max isochrone coverage}
# Driving range
drivingMAX_isos <- mb_isochrone(
  sch_sf,
  profile = "driving",
  time = 55,
  id = "label"
)

# map
map_driveMAX <- mapbox_map  %>%
  addPolygons(data =drivingMAX_isos,
             popup = ~id,
             color = colors2[2],
             fillColor = colors2[2],
             fillOpacity = 1,
             opacity = 0.3,
             weight = 2,
             highlight = highlightOptions(weight = 0.8,
                                       color = "blue",
                                       bringToFront = TRUE)) %>%
  addLegend(position = "topright",
           colors = colors2[2],
           labels = "50",
           title = "Drive reach (minutes)")

# show map
map_driveMAX

# save data
write_sf(drivingMAX_isos, "../data/w_data/driveMAX_isos4326.shp")
```


```{r Calculate area for MAX, warning=FALSE}
# same calculations as above
drivingMAX_tf <- st_transform(drivingMAX_isos, 25832)
drivingMAX_un_tf <- st_union(drivingMAX_tf)
difference3 <- st_difference(DK_tf, drivingMAX_un_tf)
st_crs(difference3)
# plot
plot(st_geometry(difference3), col = "grey", border = "white")

st_area(DK_tf) # area of the whole Denmark
st_area(difference3) # area which can't reach a special school within 55min drive
st_area(difference3)/1000000 # calculate km^2
st_area(difference3)/st_area(DK_tf) # get percentage

```

The area of Denmark not able to reach a Special school in 55 minute drive is 1358.847 km^2 which is 3.15% I am assuming the 3.15% accounts for islands without special schools and road connection to mainland. Also beaches, parks, forests etc. with no roads.

# Coverage by municipality
```{r Load municipality shapefiles}
# loading data for municipalities and transforming to same epsg
mun <- getData("GADM", country = "DNK", level = 2)
mun_tf <- st_transform(st_as_sf(mun), crs = 32618)
mun_tf <- st_transform(mun_tf, 25832)
st_crs(mun_tf)

# check with a plot
plot(st_geometry(mun_tf), col = "grey", border = "white")

```


## Walking - coverage calculation
```{r Calculate areas for each municipality separately - walking, results='hide', warning=FALSE}
# empty lists to save our values
total <- list()
ina_area <- list()
ina_percent <- list()

# loop over municipalities and calculate the same results as above, but for each munic
for (x in 1:nrow(mun_tf)) {
  print(x) # print current index to see progress
  M <- mun_tf[x, ] # extract specific municipality
  
  diff <- st_difference(M, walking_un_tf) # calculate difference
  
  # I had to create a condition because some municipalities are fully covered and the polygon of difference doesnt 
  # exist in that case (I illustrate the problem at the end of this markdown with visuals)
  if (nrow(diff) == 0) {
    a <- st_area(M)
    b <- 0
    c <- 0
  } else {
    a <- st_area(M) 
    b <- st_area(diff) 
    c <- st_area(diff)/st_area(M) 
  }
  # append results to our lists
  total <- append(total, as.integer(a))
  ina_area <- append(ina_area, as.integer(b))
  ina_percent <- append(ina_percent, as.numeric(c))
  
}

# multiply by 100 to get percentage
ina_percent <- lapply(ina_percent,"*",100)

# prepare my columns for new result df
municipality <- mun_tf$NAME_2
total_area <- unlist(total)
out_reach <- unlist(ina_area)
out_reach_perc <- unlist(ina_percent)

# make df with results
municipalities_res <- data.frame(municipality, total_area, out_reach, out_reach_perc)
# round the percentages to two decimal points
municipalities_res$out_reach_perc <- round(municipalities_res$out_reach_perc ,2)
# save the file
write_csv(municipalities_res, "../data/results/munic_walk.csv")

```


## Driving - coverage calculation
```{r Calculate areas for each municipality separately - driving, results="hide", warning=FALSE}
# This is the same code but applied to our driving data
total <- list()
ina_area <- list()
ina_percent <- list()


for (x in 1:nrow(mun_tf)) {
  print(x)
  M <- mun_tf[x, ]
  
  diff <- st_difference(M, driving_un_tf)
  
  if (nrow(diff) == 0) {
    a <- st_area(M)
    b <- 0
    c <- 0
  } else {
    a <- st_area(M) 
    b <- st_area(diff) 
    c <- st_area(diff)/st_area(M) 
  }
  
  total <- append(total, as.integer(a))
  ina_area <- append(ina_area, as.integer(b))
  ina_percent <- append(ina_percent, as.numeric(c))
  
}


ina_percent <- lapply(ina_percent,"*",100)

municipality <- mun_tf$NAME_2
total_area <- unlist(total)
out_reach <- unlist(ina_area)
out_reach_perc <- unlist(ina_percent)


municipalities_res2 <- data.frame(municipality, total_area, out_reach, out_reach_perc)
municipalities_res2$out_reach_perc <- round(municipalities_res2$out_reach_perc ,2)
write_csv(municipalities_res2, "../data/results/munic_drive.csv")

```


# How many special school students are in each municipality?
```{r Municipalities and special students}
# load data
students <- read_csv(file="../data/w_data/student_count_cl.csv", col_names = TRUE)

# rename so we can merge with municipality dataset
colnames(students) <- c('NAME_2', 'special_st', 'total_st')

# let's check if the municipality names are the same in both dataframes
orig <- sort(setdiff(students$NAME_2, mun_tf$NAME_2)) # they are not but we know which now, so we can fix it
new <- sort(setdiff(mun_tf$NAME_2, students$NAME_2))

# made a for loop to fix the names so I don't have to write the same code many times
for (x in 1:3) {
  students$NAME_2[students$NAME_2 == orig[x]] <- new[x]
  }

# join the dataframes
mun_tf_st = mun_tf %>% inner_join(students ,by="NAME_2")
mun_tf_st

# check crs
st_crs(mun_tf_st)
```

```{r map Special students in municipalities}
#max(na.omit(mun_tf_st$special_st))
# set breakpoints for map
breakpoints = seq(0, 2700, 300)

# map
mapview(mun_tf_st, 
        zcol = "special_st",
        col.regions = viridisLite::inferno,
        at = breakpoints,
        layer.name = "Special school students") +
  mapview(sch_sf,
          popup = sch_sf$label,
          cex = 2,
          color = "green",
          color.fill = "green",
          layer.name = "Special schools",
          legend = TRUE)
```
We can see that some municipalities with higher numbers of students have only 1 or 2 schools, some 0.
We could assume students travel further/longer in these areas.



## Walking - map of coverage
```{r Municipality area by walking reach}
# load the results we stored
w_access <- read_csv(file="../data/results/munic_walk.csv", col_names = TRUE)
# rename columns so we can merge them with munic data
access_walk <- w_access %>% rename(NAME_2 = municipality) %>% rename(out_reach_w = out_reach_per)
# remove unneccessary columns so the program won't crash
access_walk <- access_walk[ ,c(1,4)]
# check if kommunes have the same names
setdiff(access_walk$NAME_2, mun_tf$NAME_2)
setdiff(mun_tf$NAME_2, access_walk$NAME_2) # seems alright
# join dataframes
mun_tf_merg = mun_tf_st %>% inner_join(access_walk ,by="NAME_2")
mun_tf_merg

# set breaks at every 10%
at_lb = seq(0, 100, 10)
pal <- magma(n = length(at_lb), direction = 1)
# make a map
mapview(mun_tf_merg, zcol = "out_reach_w",
        col.regions = pal,
        at = at_lb,
        layer.name = "% is outside of 5km walking range")

```

## Driving - map of coverage
```{r Municipality area by driving reach}
# same process but with driving data
# load data
d_access <- read_csv(file="../data/results/munic_drive.csv", col_names = TRUE)
# prepare for merging
access_drive <- d_access %>% rename(NAME_2 = municipality) %>% rename(out_reach_d = out_reach_perc)
access_drive <- access_drive[ ,c(1,4)]
# check names
setdiff(access_drive$NAME_2, mun_tf$NAME_2)
setdiff(mun_tf$NAME_2, access_drive$NAME_2)
# merge
mun_tf_merg_full = mun_tf_merg %>% inner_join(access_drive ,by="NAME_2")
mun_tf_merg_full

# set breakpoints
at_lb = seq(0, 100, 10)
pal <- magma(n = length(at_lb), direction = 1)
# map
mapview(mun_tf_merg_full, zcol = "out_reach_d",
        col.regions = pal,
        at = at_lb,
        layer.name = "% is outside of 30 min drive range")

```

# Students in areas with low accessibility 
```{r Calculating how many students live in low accessibility areas}
# filter all the rows with the percentage column higher or equal to 75
lowest_w <- mun_tf_merg_full %>%
  filter(out_reach_w >= 75)
# add up all the values from the student count column and ignore NAs
sum(na.omit(lowest_w$special_st))
# 22327 students live in municipalities with 75% of area outside of range of 5km walk

# same calculation but with 90%
lowest_w2 <- mun_tf_merg_full %>%
  filter(out_reach_w >= 90)

sum(na.omit(lowest_w2$special_st))
# 17044 students live in municipalities with 90% of area outside of range of 5km walk

lowest_w3 <- mun_tf_merg_full %>%
  filter(out_reach_w >= 100)

sum(na.omit(lowest_w3$special_st))
# 2643 students live in municipalities with 100% of area outside of range of 5km walk

# driving - same calculation
lowest_d <- mun_tf_merg_full %>%
  filter(out_reach_d >= 50)

sum(na.omit(lowest_d$special_st))
# 663 students live in municipalities with 50% of area outside of range of 30 minute drive

lowest_d2 <- mun_tf_merg_full %>%
  filter(out_reach_d >= 75)

sum(na.omit(lowest_d2$special_st))
# 181 students live in municipalities with 75% of area outside of range of 30 minute drive

```


## Illustrative example
- here I include a visualisation of some municipalities completely overlapped by our access polygons and why the difference polygon can't be made
```{r, explanatory example, warning=FALSE}
# let's check what is 10th kommune in our df
mun_tf$NAME_2[10]
# we extract only "Frederiksberg" so we can work with it
M <- mun_tf[10,]
# plot to check if it looks okay
plot(st_geometry(M), col = "grey", border = "white") # it looks fine so let's continue

# calculate the area of difference between our walking polygon and the kommune
diff <- st_difference(M, walking_un_tf)
# check it out
diff # Hmm, it's empty! Why would that be? 

# Well, perhaps the polygon representing the difference can't be made because the difference does not exist

# Here I am plotting the two on one map to see if my assumption is correct
  
mapFrederiksberg <- mapbox_map  %>%
  addPolygons(data =walking_isos,
             popup = ~id,
             color = colors2[2],
             fillColor = colors2[2],
             fillOpacity = 0.2, 
             opacity = 0.3,
             weight = 2, 
             highlight = highlightOptions(weight = 1,
                                       color = "blue",
                                       bringToFront = TRUE)) %>%
  addPolygons(data = mun[10,],
              color = "red",
              fillOpacity = 1
              )
mapFrederiksberg

# indeed, if we zoom in on the red area, it' s completely covered by different polygons we created with walking 
# distances

```